---
title: "Demographic Data Summaries with `zoopop`"
tutorial:
  id: "demography"
  version: 1.0
author: "Alicia Rich"
date: "`r Sys.Date()`"
output: learnr::tutorial
runtime: shiny_prerendered
                     
---

## Introduction

### Goal

This tutorial will walk through example data summary and visualization tools using the `zoopop` package.

### Details

I am going to use the Pygmy Slow Loris population reports to generate a reproducible report that summarizes:

1.  Population Size and Composition

2.  Historic Demographic Trends Compared Across Generations

2.  Contemporary Patterns of Genetic Diversity and Relatedness

3.  Projected Population Viability

### Files Needed

I am loading two csv files created with the `zoopop` package: 

1.  A studbook organized into rows containing static profiles of individuals across the entire population history and individual transfer and breeding events over each individual's lifetime: `studbook.csv`

2.  A table where every institution appearing in the studbook is represented by a single row: `locations.csv`

For more information, see the tutorial below.

```{r eval=FALSE}
learnr::run_tutorial("new_studbook", package = "zoopop")
```



## Setup

```{r setup, context="setup"}
here::i_am("inst/tutorials/demography/demography.Rmd")
library(conflicted)
library(here)
library(knitr)
library(learnr)
library(shiny)
library(tidyverse)
library(zoopop)
knitr::opts_chunk$set(
  collapse = TRUE,
  message  = FALSE,
  warning  = FALSE,
  echo     = TRUE,
  include  = TRUE,
  eval     = TRUE,
  comment  = ""
)
conflicts_prefer(lubridate::interval)
conflicts_prefer(dplyr::filter)
loc_key  <- read_csv(here("inst/extdata/locations.csv"))
studbook <- read_csv(here("inst/extdata/studbook.csv"))
```


```{r setup, context="setup"}
cols_default <- c(
  "Status"              ,
  "ID"                  ,
  "name_spec"           ,
  "exclude"             ,
  "Loc_birth"           ,
  "Loc_last"            ,
  "Sire"                ,
  "Dam"                 ,
  "Date_birth"          ,
  "age_last"            ,
  "Date_last"           ,
  "Sex"                 ,
  "Institution_birth"   ,
  "iconLoc_birth"       ,
  "Institution_last"    ,
  "iconLoc_last"        
)

```

# Studbook Overview

Let's create a graphical table to summarize our studbook data.

```{r studbook-view, exercise=TRUE}
studbook.tbl <- studbook %>% studbook_react(., df_cols = c(cols_default), groupBy = "Status")

studbook.tbl
```

## Summarize Demographic Trends

Let's look at the basic population size over time to begin quantifying growth rate and assessing overall population viability.

```{r census, exercise=TRUE}
population.plot <- plot_census(census(studbook, "years"), title = "Pygmy Loris Population Trends 1966-2025")
population.plot
```

## Life Table Calculations 

Now we will use annual birth counts to follow the standard method for computing a normalized rate of growth over the past 5 years compared to all years.

```{r lifetab, exercise = TRUE}
lifeTab      <- cohort_lifeTab(studbook, 
                               Year_min = 1985, 
                               Year_max = 2024, 
                               span     = 5)
```

And then we will plot trends in this growth rate based on birth year.

```{r lambda, exercise = TRUE, exercise.setup = "lifetab"}
lambda.plot  <- plot_lambda(lifeTab, 
                            title = "Population Growth Rates by Cohort")
lambda.plot
```

## Construct Pedigree 

Now we will use `pedtools` to create our pedigree objects based on the studbook data. We will use the pedigree object to generate all genetic stats and projections.

```{r pedigree, exercise = TRUE}
ped_series <- build_ped_series(studbook)
```

When working with studbook data, this usually produces a list of multiple pedigree objects, as any founders or lineages not represented in the current living population will form discrete, disconnected trees. There also will probably be several "singletons" which are founders without any connections to the living population. That is why the code above also included a line to thin out your dataset by ignoring the founders that never reproduced.

### Pedigree Plot

Now we will take a look at the data in the form of the traditional pedigree structures produced by `pedtools`. The code below also prepares some formatting variables to assign color-coding that matches the other graphics in this workflow before plotting the pedigree(s).

```{r pedplots, exercise = TRUE, exercise.setup = "pedigree"}
ped_series_plots <- plot_ped_series(ped_series, 
                                    studbook, 
                                    labs_spec = TRUE)
ped_series_plots
```

We can download this entire series to a local directory as html files using the function below.

```{r}
pedsave_series(ped_series, studbook, labs_spec = TRUE)
```

For many of our operations, it will be easier to work with one pedigree object of interest and ignore the pedigrees not currently represented in the breeding population. The function below helps you extract the pedigree of interest as its own object with a new name.

```{r}
ped <- extract_pedigree(ped_series, "_comp1", "Contemporary")
```

Now we can plot just this pedigree on its own.

```{r}
ped_plot <- plot_pedigree(ped, studbook, labs_spec = TRUE)
ped_plot
```

And to download this as a single html file we can use the function below.

```{r}
pedsave(ped, studbook, labs_spec = TRUE)
```

#### Interactive Pedigree Graphics

We can also adapt `visNetwork` functions to render javascript-based graphics to represent the pedigree relationships built by `pedtools.`

```{r}
nodes <- ped_ndf(ped, studbook)
```


